\bf Distortion corrected kernel density estimator on Riemannian manifolds

### 2024-09-11 jcgs/03-chap3-kderm.Rmd
Removed colors in text

One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$. To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead. To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors. We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density. The entire workflow is summarized in \autoref{fig:vkde}. The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection. Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted in blue shades. With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.

@@

```{r tpoutliers, fig.pos = "!p", fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row. The top 20 outliers, highlighted in black and indexed in blue text, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel). DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
```

\autoref{fig:tpoutliers} summarizes the results. Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively. We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$. Colors show the different estimated densities at each point with anomalies shown in darker colors with blue indexing text, and higher density points shown in lighter colors. For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions. On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.  The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies. For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes. This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data [@cai2022-tsne].

@@

In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction. Data are presented on a log scale to highlight anomalies. The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a $99\%$ highest density region. The middle shaded region contains anomalies that are true positives in the sense of not lying in a $90\%$ HDR. Points lying outside squares (shown as triangles) are incorrectly classified. For example, the three red triangles in the middle left of the left panel are truly anomalies since they lie outside the $99\%$ HDR, but are not classified as such (although they are true positives if a 90% HDR is used). Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection. \autoref{fig:tptsneden} shows the same plot but for t-SNE. The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified. However, the difference between KDE with and without distortion correction is stark. These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.

@@

```{r electricityembedding, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel). The black points with blue text indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
knitr::include_graphics("figures/Electricity_2d_5levels_outliers_comparison_5ml_r180.png")
```
```{r electricityoutliers, eval=T}
load("figures/Electricity_2d_table_outlier_idcompare_5ml_radius20_r180.rda")
# tabletop20 <-
  f20 %>%
  kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding. The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE. The blue-colored IDs are IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
                  col.names = c(" ", rep(c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"), 2))
                  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  kable_paper(full_width = TRUE) %>%
  # add_header_above(c(" " = 1, rep(c("ISOMAP" = 1, "LLE" = 1, "Laplacian Eigenmaps" = 1, "t-SNE" = 1, "UMAP" = 1), 2))) %>%
  add_header_above(c(" " = 1, "DC-KDE" = 5, "KDE" = 5)) %>%
  # row_spec(1, bold = TRUE) %>%
  column_spec(1, bold = FALSE) %>%
  column_spec(2, bold = freqmax[,2], color = f20cols[,2]) %>%
  column_spec(3, bold = freqmax[,3], color = f20cols[,3]) %>%
  column_spec(4, bold = freqmax[,4], color = f20cols[,4]) %>%
  column_spec(5, bold = freqmax[,5], color = f20cols[,5]) %>%
  column_spec(6, bold = freqmax[,6], color = f20cols[,6]) %>%
  column_spec(7, bold = freqmax[,7], color = f20cols[,7], border_left = TRUE) %>%
  column_spec(8, bold = freqmax[,8], color = f20cols[,8]) %>%
  column_spec(9, bold = freqmax[,9], color = f20cols[,9]) %>%
  column_spec(10, bold = freqmax[,10], color = f20cols[,10]) %>%
  column_spec(11, bold = freqmax[,11], color = f20cols[,11])
# tabletop20
```

\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE. There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions. The dark-colored points indexed with blue household IDs are the top 20 anomalies with the lowest densities. First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings. When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE. As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies. The darker blue points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter blue points are those identified four times. Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases. Without distortion correction, the density estimates are different not only in scale, but also show inconsistency in the rank of density estimates.

@@

Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households. To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) where anomalies fall outside the $>99\%$ and $99\%$ HDRs respectively that are also identified as anomalies by different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP. \autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods. The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}. The household indices shown in darker shades of blue are identified as anomalies the most with dimension reduction algorithms. It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.

>>>>

One limitation of the kernel density estimator is that the density can be estimated only at points where data have been observed since the estimator requires the Riemannian $\bm{H}_j$. To estimate the density at points that do not correspond to observed data, any smoothed average of nearest neighbors can be used instead. To account for distortion, this average should be weighted by the determinant of the Riemannian at each of the nearest neighbors. We note that the particular downstream task that we are interested in is anomaly detection for which only the density estimates at observed sample points are required since anomalies are identified as the points with the lowest density. The entire workflow is summarized in \autoref{fig:vkde}. The last two steps in \autoref{fig:vkde} are our main contributions, generating distortion-corrected KDE with adaptive Riemannian metric $\bm{H}_i$ at each point and computing the highest density region plots based on the density estimates for anomaly detection. Compared to the anomaly detection with a general kernel density estimator in @Cheng2021-dh, the changes are also highlighted with shading. With this anomaly detection process, outliers based on the lowest densities could be detected more accurately regardless of the distortion in manifold learning.

@@

```{r tpoutliers, fig.pos = "!p", fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of five manifold learning embeddings of the twin peaks data in each row. The top 20 outliers, highlighted as points and indexed with text labels, are found by the true manifold density (left panel), DC-KDE (middle panel), and KDE (right panel). DC-KDE finds more true outliers than KDE in all five rows.", echo = FALSE}
knitr::include_graphics("figures/Twin Peak2000_5levels_outliers_comparison_5ml_r0_5.png")
```

\autoref{fig:tpoutliers} summarizes the results. Each row of panels corresponds to a different dimension reduction technique, while the left, center and right columns correspond to density estimates for the ground truth density, distortion-corrected KDE, and KDE respectively. We set the bandwidth parameter in the DC-KDE \@ref(eq:denestimator) as $r=0.5$. The color shades show the different estimated densities at each point with anomalies shown in darker colors with indexing labels, and higher density points shown in lighter colors. For many methods, the salient features of the ground truth distribution are clear regardless of whether distortion correction is applied, for example for ISOMAP, all three plots, identify a similar set of outliers and four high-density regions. On the other hand for LLE, the left panel shows that dimension reduction pulls outliers on the manifold towards the center.  The distortion-corrected KDE can account for this, while KDE without distortion correction on the other hand does not correctly identify the anomalies. For t-SNE, the ground truth and distortion-corrected KDE identify four regions of high density, while a KDE estimate without distortion correction seems to identify a larger number of modes. This concurs with the common observation that t-SNE tends to output clusters even where such clusters may not be present in the underlying data [@cai2022-tsne].

@@

In \autoref{fig:tpisomapden}, we plot ranks of the estimated density against the true density for the ISOMAP embedding with the left panels showing results for distortion correction and the right panel showing results without distortion correction. Data are presented on a log scale to highlight anomalies. The bottom left shaded region contains all points that are truly anomalies and are identified as such (true positives), where an anomaly is defined as a point not falling within a $99\%$ highest density region. The middle shaded region contains anomalies that are true positives in the sense of not lying in a $90\%$ HDR. Points lying outside squares (shown as triangles) are incorrectly classified. For example, the four colored triangle points in the middle left of the left panel are truly anomalies since they lie outside the $99\%$ HDR, but are not classified as such (although they are true positives if a 90% HDR is used). Overall, the right panel contains many more misclassified anomalies, which shows that failing to apply distortion correction can have a severe impact on anomaly detection. \autoref{fig:tptsneden} shows the same plot but for t-SNE. The quality of t-SNE is worse than ISOMAP in this example therefore many more anomalies are misclassified. However, the difference between KDE with and without distortion correction is stark. These results highlight the importance of applying distortion correction especially when the quality of dimension reduction may not be high.

@@

```{r electricityembedding, fig.align = 'center', out.width = "85%", fig.cap="Highest density region plots of the 2-D embedding for $3,639$ households, with each point representing the distribution of one household and colors indicating the density estimated from DC-KDE (left panel) and KDE (right panel). The highlighted points with ID indexing are the top 20 anomalous households found by each dimension reduction method in each row.", echo=FALSE, eval=T}
knitr::include_graphics("figures/Electricity_2d_5levels_outliers_comparison_5ml_r180.png")
```
```{r electricityoutliers, eval=T}
load("figures/Electricity_2d_table_outlier_idcompare_5ml_radius20_r180.rda")
# tabletop20 <-
  f20 %>%
  kableExtra::kbl(caption = "Top 20 anomalous household IDs using density estimates of different 2-dimensional embedding. The left five columns are anomalies by DC-KDE, while the right five columns are anomalies using KDE. The colored IDs are household IDs that are detected over three times, with the darker one indicating five times and the lighter one indicating four times.", booktabs = TRUE, digits = 3, escape = FALSE, align=rep('c', 10),
                  col.names = c(" ", rep(c("ISOMAP", "LLE", "Laplacian Eigenmaps", "t-SNE", "UMAP"), 2))
                  ) %>%
  kable_styling(latex_options = "scale_down") %>%
  kable_paper(full_width = TRUE) %>%
  # add_header_above(c(" " = 1, rep(c("ISOMAP" = 1, "LLE" = 1, "Laplacian Eigenmaps" = 1, "t-SNE" = 1, "UMAP" = 1), 2))) %>%
  add_header_above(c(" " = 1, "DC-KDE" = 5, "KDE" = 5)) %>%
  # row_spec(1, bold = TRUE) %>%
  column_spec(1, bold = FALSE) %>%
  column_spec(2, bold = freqmax[,2], color = f20cols[,2]) %>%
  column_spec(3, bold = freqmax[,3], color = f20cols[,3]) %>%
  column_spec(4, bold = freqmax[,4], color = f20cols[,4]) %>%
  column_spec(5, bold = freqmax[,5], color = f20cols[,5]) %>%
  column_spec(6, bold = freqmax[,6], color = f20cols[,6]) %>%
  column_spec(7, bold = freqmax[,7], color = f20cols[,7], border_left = TRUE) %>%
  column_spec(8, bold = freqmax[,8], color = f20cols[,8]) %>%
  column_spec(9, bold = freqmax[,9], color = f20cols[,9]) %>%
  column_spec(10, bold = freqmax[,10], color = f20cols[,10]) %>%
  column_spec(11, bold = freqmax[,11], color = f20cols[,11])
# tabletop20
```

\autoref{fig:electricityembedding} presents the highest density regions of the 2-dimensional embedding for each dimension reduction method, ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP, with the left panels showing the densities estimated with the distortion corrected KDE and the right panels showing those estimate with KDE. There are six density regions colored in the plot, $1\%$, $5\%$, $50\%$, $90\%$, $99\%$ and $>99\%$, and darker colors indicate lower density regions. The dark-colored points indexed with household IDs are the top 20 anomalies with the lowest densities. First of all, for the same electricity data, each dimension reduction method generates quite different embeddings, meaning that distortions exist in these embeddings. When comparing row-wise for the same embedding, it can be seen that ISOMAP and LLE give similar anomalies while the outliers are more clustered for Laplacian Eigenmaps, t-SNE, and UMAP with DC-KDE compared with the KDE. As the anomalous IDs are not easy to distinguish in the plots, we list them in \autoref{tab:electricityoutliers} and highlight the IDs that are most identified as anomalies. The darker colored points are those households that are identified five times out of the ten cases shown in \autoref{fig:electricityembedding}, and lighter colored points are those identified four times. Again, KDE with distortion correction methods finds more anomalies in common regardless of the dimension reduction methods, compared with KDE where only LLE could find similar anomalies to other cases. Without distortion correction, the density estimates are different not only in scale, but also show inconsistency in the rank of density estimates.

@@

Unlike for the simulations where the ground truth densities are known, in the applications, we have no knowledge of the truly anomalous households. To explore the robustness of anomaly detection both with and without distortion correction, we assume, in this case, that the anomalies given by UMAP are the `true' ones and evaluate the performance of the other four dimension reduction methods, ISOMAP, LLE, Laplacian Eigenmaps and t-SNE. The results are summarized in \autoref{tab:electricityhdrpercent} which can be interpreted similarly to \autoref{tab:fourdhdrtable}. \autoref{tab:electricityhdrpercent} presents the proportions of 'true' outliers (given by UMAP) where anomalies fall outside the $>99\%$ and $99\%$ HDRs respectively that are also identified as anomalies by different dimension reduction methods. Both DC-KDE and KDE are shown in the table. Apart from the lower percentage of ISOMAP in $>99\%$ HDRs, distortion-corrected KDE gives higher proportions of true outliers in all other methods. The improvement brought by distortion correction is much more pronounced in Laplacian Eigenmaps and t-SNE which would otherwise detect almost completely different anomalies to UMAP. \autoref{fig:electricityrank} presents the visualization of density ranking in log scale. The grey shading area covers the density rank from 1 to 20, which are the anomalies listed in \autoref{tab:electricityoutliers}. Again, the plots show the robustness of density estimated with distortion correction regardless of the dimension reduction methods. The robustness of anomaly detection to different methods when using distortion correction is also shown in \autoref{tab:electricityoutliers}. The household indices shown in darker shades are identified as anomalies the most with dimension reduction algorithms. It is notable that without applying distortion correction, Laplacian Eigenmaps, t-SNE and UMAP do not identify any of the same anomalies as ISOMAP and LLE.
